---
title: "Lab6"
author: "Mustafa Cavus"
date: "2025-11-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# loading the necessary packages
install.packages("ggplot2")
install.packages("dplyr")
library(ggplot2)
library(dplyr)
```

# Dataset

For detailed information about the dataset, check here:
https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars

```{r}
# loading the mtcars dataset
data("mtcars")
head(mtcars)
```

# Drawing a scatter plot

Let's visualize the association between horsepower (`hp`) and miles per gallon (`mpg`) variables in the mtcars dataset. To do this, we can use either the `plot()` function or the `{ggplot2}` package.

```{r}
plot(mtcars$hp, mtcars$mpg,
     xlab = "hp",
     ylab = "mpg",
     main = "Visualization of the association between mpg and hp")
```

Visualization using `{ggplot2}`:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp")
```


Let's also examine this association by considering the transmission type of the cars. This way, we can see if the association between these two variables differs depending on the transmission type (automatic or manual):

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, colour = am)) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp")
```

When you use the above commands, you'll notice that the transmission type (`am`) variable, which has only two values, is visualized as if it has five different values between 0 and 1. The reason for this is that the transmission type variable is not defined as a categorical variable. In such cases, we can use the `factor()` function to convert the variable into a categorical one.

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, colour = factor(am))) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp")
```

When you run the above commands, you'll see that the labels for the color aesthetics representing the transmission type are 0 and 1. To make this more understandable, we need to identify whether the 0 and 1 labels represent automatic or manual transmission. For this, we can refer to the link provided at the top of this document.

```{r}
ggplot(mtcars, aes(x = hp, y = mpg, colour = factor(am))) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp by transmission type",
       colour = "Transmission Type") + 
  scale_colour_discrete(labels = c("Automatic", "Manual"))
```

To create the same visualization in a side-by-side layout, we need to use the `facet_grid` feature:

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp by transmission type") + 
  facet_grid(~am)
```

When you run the above commands, you'll notice that the labels for each grid created for the transmission type need adjustment. Therefore, the 0 and 1 values of the transmission type variable should be changed to "Automatic" and "Manual". The following code can be used for this transformation:

```{r}
mtcars$am <- ifelse(mtcars$am == 1, "Manual", "Automatic")
```

Then,

```{r}
ggplot(mtcars, aes(x = hp, y = mpg)) + 
  geom_point() + 
  labs(title = "Visualization of the association between mpg and hp by transmission type") + 
  facet_grid(~am)
```

In almost all analyses, the associations between all continuous variables in the dataset are explored. The most useful tool for this purpose is a scatter matrix. First, the continuous variables in the dataset need to be identified.

After checking the detailed information about the dataset, we find that mpg, disp, hp, drat, and wt are continuous variables. A scatter matrix can be drawn by selecting only the columns containing these variables:

```{r}
plot(mtcars[, c(1,3:6)])
plot(mtcars[, c(1,3)])
```

# Drawing a correlogram

The GGally, corrgram, and ellipse packages provide functions for drawing correlograms. For detailed information, check here: 
https://www.r-graph-gallery.com/correlogram.html

```{r message=FALSE, warning=FALSE, include=FALSE}
install.packages("GGally")
library("GGally")
```

Then,

```{r}
ggcorr(mtcars[, c(1,3:6)]) 
```

To make the correlogram more readable, we can use shape aesthetics. For this, we can use the `plotcorr()` function from the `{ellipse}` package. First, we need to calculate the correlation coefficients of the variables in the dataset using the `cor()` function. Let's assign this to the `cor_mtcars` object. After that, we can pass this object to the `plotcorr()` function to generate the visual:

```{r message=FALSE, warning=FALSE, include=FALSE}
install.packages("ellipse")
library(ellipse)
```

Then,

```{r}
cor_mtcars <- cor(mtcars[, c(1,3:6)])
plotcorr(cor_mtcars[order(cor_mtcars[,1]), order(cor_mtcars[,1])])
```
